A Case Study on Abstractions to Converge High-Throughput and High-Performance Computing

### 2017-03-19 PandaOverview.tex
Ongoing iteration III, preparing for new diagram

The design of the Production and Distributed Analysis (PanDA) system started in
2005 to develop a WMS for the ATLAS experiment. PanDA went into production for
the LHC Run 1 on 2009 and was then partially redesigned to be deployed on Run 2,
on 2015. In this section, we summarize the design, architecture, and execution
process of the current state of the practice.

@@

PanDA's application model primarily assumes single tasks, workloads and
workflows. Tasks represent a set of operations performed on a set of events,
stored across input files. Tasks can be decomposed into jobs, where each job
represents the task's set of operations and a partition of the task's events.
Since 2005, a certain amount of parallelism has been progressively introduced
for jobs~\cite{crooks2012multi} but, so far, no MPI jobs have been considered
for production. Jobs are supposed to be relatively self-contained, capable of
setting up their own execution environment or having a minimal set of common
dependences. Different types of job payload (i.e., executable) are assumed, both
in terms of user-defined batch scripts or application frameworks for group of
users.

PanDA's user model is based on multitenancy and the support of at least two
types of HEP users: individual researchers and groups executing production-grade
workflows. Users are considered free to submit tasks and workflows to the PanDA
WMS at any point in time, directly or via dedicated application frameworks.
Consistently, PanDA's security model is based on separation between
authentication, authorization and accounting for both single users and group of
users. Both authentication and authorization are based on digital certificates
and the X.509 standard and on the virtual organization abstraction.

Currently, PanDA's execution model is based on five main abstractions: task,
job, queue, pilot, and event. Both tasks and jobs are assumed to have attributes
and states and to be queued into a global queue for execution. Prioritization
and binding of jobs are assumed to depend on the attributes of each job. Pilot
is used to indicate the abstraction of resource capabilities. Jobs are thought
to be bound to pilots and the execution of their payload is performed on the
site where the pilots has been instantiated. Consistently, pilots are considered
to be multitenant as jobs belonging to multiple workflows and/or users can be
bound and executed on them.

PanDA' data model uses events as the unit of data. Each event represents a
collision captured in the detector and one or more events can be packaged into
files or data source with other container abstractions. As with jobs, data have
both attributes and states, and some of the attributes are shared between events
and jobs so to create associations. Raw, reconstruction, and simulation data are
assumed to be distributed across multiple and tiered sites and managed by the
ATLAS Distributed Data Management (DDM)~\cite{garonne2012atlas}. Datasets
required by each job are assumed to be replicated when necessary over the
network. This applies to both input and output data. \sergeynote{Raw data from
the ATLAS detector is captured at CERN and goes through reconstruction step to
produce data in secondary formats, smaller in volume and with more and more
physics content instead of raw detector hits. Consequently a second, custodial,
copy of raw data is being distributed in predetermined volumes to National T1
centers around the world (like BNL  T1 center). This insures low raw data
redundancy and safety. Multiple copies of chunks of secondary data sets (derived
from real raw data) are also distributed to ~100 centers around the word for
ease of access and improved availability for analysis. Then raw detector data is
stored on Tapes at CERN, just  in case. Most users, who do physics analysis,
rarely work with raw data, only a small group of experts needs to work with raw
data for specialized tasks like detector calibrations and such. There may be a
need to redo a reconstruction step, when errors in previous reconstruction
campaign are uncovered or improved detector understanding will lead for a better
quality physics. Then data stored on tape is recovered and re-processed again.
Simulation data is produced in many places anyway and is not centralized to
begin with. Here's the paper that describes current DDM state in ATLAS, data
replication strategies and such.
http://iopscience.iop.org/article/10.1088/1742-6596/396/3/032045/pdf DDM is a
complex and evolving topic in ATLAS.}\mtnote{Thank you. As we are describing the
design of PanDA I thought we should not enter into a too detailed description of
the DDM design. I eliminated the mistake I had made about `centralized data' but
left DDM just as a reference. Any better?}

PanDA design supports provenance and traceability for both jobs and data.
Attributes enable provenance by linking jobs and data items, providing
information like ownership or project affiliation. States enable traceability by
providing information about the stage of the execution in which each job or data
item is or have been. Some attributes are assumed to be immutable across
execution and jobs and data items are assumed to be always in one and only one
state.

@@

The implementation of PanDA consists of several interconnected subsystems, most
of them built from off-the-shelf and Open Source components. At a high level,
PanDA's architecture has six main subsystems: PanDA Server, AutoPyFactory, PanDA
Pilot, PanDA Monitoring, JEDI, and Event Service. Subsystems communicate via
dedicated API or http messaging. Databases are used to store eventful entities
like tasks, jobs and events, and to store information about resources, logs, and
accounting. Each subsystems is implemented by one or more modules as shown in
Figure~\ref{fig:architecture}. In the following, we briefly describe the overall
architecture of PanDA and its components. The description provides all the
details required to understand how HPC resources have been integrated within
PanDA (\ref{sec:integration}).

@@

\paragraph{\textbf{PanDA Server}} this subsystem is the central hub service of
PanDA and it operates as a web service. It runs on Apache web server,
interacting with the back-end database running on separate servers. It
implements the Apache worker model, with many independent processes handling
client requests in parallel. Since all state is maintained in the central
database, the PanDA Server application instance itself is stateless. Currently
production version of PanDA utilizes Oracle database backend, but PanDA can also
work with MySQL family of databases.

@@

PanDA Server implementation consists of four components, all implemented in
Python and communicating via Python APIs: Task Buffer, Brokerage, Job
Dispatcher, and Data Service. Task Buffer is the management component of PanDA's
global queue and offers functionalities to: (i) store jobs into the global
queue; (ii) provide information about and update the state of each stored job.
The Brokerage component enables (i) matching job attributes with site and pilot
attributes; (ii) managing the dispatch of input data to processing sites; and
(iii) deriving PanDA's data pre-placement requirement.
% The job/pilot matching algorithm assigns a weight to each candidate pilot and
% after applying a set of policies, the best ten candidates are taken.
The Job Dispatcher implements functionalities to: (i) receive requests from
active pilots for new jobs to process; (ii) retrieving from Task Buffer the jobs
with highest priority that can be executed on the resource on which the
requesting pilots have been instantiated; (iii) creating a wrapper for the
selected job, tailored to the resource of the requesting pilot; and (iv)
monitoring the state of the job execution on the pilot, updating the Task Buffer
component. Finally, the Data Service component offers functionalities to: (i)
retrieve information from Task Buffer and Brokerage about the data requirements
of each job and the resource to which each job has been assigned; (ii) enable
interaction with the ATLAS DDM to make input files available to the jobs and
manage their output data.

\paragraph{\textbf{PanDA Pilot}} this subsystem implements the execution
environment for PanDA's jobs. Implemented in Python, a PanDA Pilot is launched
by a wrapper submitted to a site. Wrappers are tailored to the
specific interfaces and middleware exposed by each site so that PanDA Pilot
remains site-independent. PanDA Pilot implements four main functionalities: (i)
preparing the execution environment for PanDA's jobs,
% including  recovering and cleaning up previously failed jobs on that site and
collecting information about the compute, data, and network capabilities of the
site; (ii) requesting a job for execution to the Job Dispatcher component of
PanDA Server; (iii) initiating and monitoring the execution of pulled jobs,
including setting up the job, transferring its input when required, executing
the payload, transferring output, and checking log files and local space; and
(iv) clean up when the payload has finished to execute.

@@

PanDA Pilot implements the pilot paradigm~\cite{turilli2015comprehensive},
decoupling resource acquisition from job execution via multi-stage scheduling.
This has several benefits, including increasing throughput by lowering the time
taken to queue and schedule jobs; enabling both concurrent and sequential job
execution depending on the number of cores available to the pilot and the time
for which it remains available; and exposing a unified and consistent interface
for the scheduling of jobs, independent from the middleware of the site on which
the pilot has been instantiated.

@@

It should be noted that upon bootstrapping, a PanDA Pilot download major parts
of its runtime code via HTTP from a central Subversion repository. The
repository works with an Apache web server, configured with a memory-based web
proxy (Squid). The purpose of the cache is to reduce the request load on the
back-end Subversion server. That improves performance \mtnote{Do we have
data/numbers?} since only occasional queries trigger a full lookup on the
back-end Subversion system, and most external queries are pulled from memory on
the front-end web server. The benefits of this system include a high-performance
code download service, combined with code updates still being immediately
available as soon as they are committed to source code control.

\paragraph{\textbf{AutoPyFactory}} this subsystem implements a factory to submit
pilots locally and to both grid and cloud
sites~\cite{caballero2012autopyfactory}. Implemented to be executed as a single
demonized process, AutoPyFactory enables the concurrent spawning of multiple job
submission processes called APFQueue. Each APFQueue can communicate with PanDA
Server and the Brokerage module to collect information about the state of the
jobs it manages and, on the basis of this information, submit new pilots to one
site per APFQueue. AutoPyFactory implements functionalities to evaluate the
number of pilots already instantiated on each site, and to apply a consistent
and adequate pressure on their batch queues. Submission is performed via the
Condor-G batch system~\cite{frey2002condor}, enabling submission to multiple
type of sites. AutoPyFactory exposes a dedicated monitoring service, feeding
information to the PanDA Monitoring subsystem. A proxy manager is also run,
enabling certificate-based authentication and authorization.

@@

\paragraph{\textbf{JEDI}} this subsystem implements a workload and workflow
management for tasks instead of jobs, as defined by HEP
users~\cite{borodin2015scaling}. Implemented as a standalone service that
interacts with Panda Server via its REST interface, JEDI offers two main
functionalities: partitioning tasks to jobs, and binding jobs to resources via
dedicated queues. JEDI partition tasks to jobs by matching tasks requirements to
resource capabilities. This matching is based on tasks attributes, and static
and dynamic information about the resources of each available site. Dynamic
information is acquired via a small number of `scout' jobs that collect
information about current resources and their performance. Functionally, JEDI
defines a set of queues associated to groups of resources with specific
capabilities. Tasks are assigned to groups of these queues, and the jobs in
which each task is partitioned are assigned to each queue, depending on their
requirements. JEDI replaces the job brokerage functionalities of the Brokerage
module of the PanDA Server but this module still handles communication between
the Task Buffer and the Data Service modules.
%, using it to enact the execution of the jobs defined by JEDI
% on the resources and sites decided by JEDI.

>>>>

The design of the PanDA WMS started in 2005 and its implementation went into
production for the LHC Run 1 on 2009. PanDA was then partially redesigned to be
deployed on Run 2, on 2015. In this section, we summarize the design,
architecture, and execution process of the current state of the practice.

@@

PanDA's application model assumes single tasks, workloads and workflows. Tasks
represent a set of operations performed on a set of events, stored in one or
more input files. Tasks are decomposed into jobs, where each job represents the
task's set of operations and a partition of the task's events. Since 2005, a
certain amount of parallelism has been progressively introduced for job
execution~\cite{crooks2012multi} but, so far, no MPI jobs have been considered
for production. Jobs are supposed to be relatively self-contained, capable of
setting up their own execution environment or having a minimal set of common
dependences.

% Different types of job payload (i.e., executable) are assumed, both in terms
% of user-defined batch scripts or application frameworks for group of users.

PanDA's user model is based on multitenancy and the support of at least two
types of HEP users: individual researchers and groups executing so called
`production' workflows. Users are free to submit tasks and workflows to the
PanDA WMS at any point in time, directly or via dedicated application
frameworks. Consistently, PanDA's security model is based on separation between
authentication, authorization and accounting for both single users and group of
users. Both authentication and authorization are based on digital certificates
and the X.509 standard and on the virtual organization (VO) abstraction.

Currently, PanDA's execution model is based on five main abstractions: task,
job, queue, pilot, and event. Both tasks and jobs are assumed to have attributes
and states and to be queued into a global queue for execution. Prioritization
and binding of jobs are assumed to depend on the attributes of each job. Pilot
is used to indicate the abstraction of resource capabilities. Each job is
thought to be bound to one pilot and executed on the site where the pilot has
been instantiated.

% Consistently, pilots are considered to be multitenant as jobs belonging to
% multiple workflows and/or users can be bound and executed on them.

PanDA' data model uses events as the unit of data\mtnote{Is event an abstraction
specific to the MD workflow?}. Each event represents a collision captured in the
detector and one or more events can be packaged into files or data source with
other container abstractions. As with jobs, data have both attributes and
states, and some of the attributes are shared between events and jobs so to
create associations. Raw, reconstruction, and simulation data are assumed to be
distributed across multiple storage facilities and managed by the ATLAS
Distributed Data Management (DDM)~\cite{garonne2012atlas}. Datasets required by
each job are assumed to be replicated when necessary over the network, both for
input and output data.

% \sergeynote{Raw data from the ATLAS detector is captured at CERN and goes
% through reconstruction step to produce data in secondary formats, smaller in
% volume and with more and more physics content instead of raw detector hits.
% Consequently a second, custodial, copy of raw data is being distributed in
% predetermined volumes to National T1 centers around the world (like BNL  T1
% center). This insures low raw data redundancy and safety. Multiple copies of
% chunks of secondary data sets (derived from real raw data) are also
% distributed to ~100 centers around the word for ease of access and improved
% availability for analysis. Then raw detector data is stored on Tapes at CERN,
% just  in case. Most users, who do physics analysis, rarely work with raw
% data, only a small group of experts needs to work with raw data for
% specialized tasks like detector calibrations and such. There may be a need to
% redo a reconstruction step, when errors in previous reconstruction campaign
% are uncovered or improved detector understanding will lead for a better
% quality physics. Then data stored on tape is recovered and re-processed
% again. Simulation data is produced in many places anyway and is not
% centralized to begin with. Here's the paper that describes current DDM state
% in ATLAS, data replication strategies and such.
% http://iopscience.iop.org/article/10.1088/1742-6596/396/3/032045/pdf DDM is a
% complex and evolving topic in ATLAS.}\mtnote{Thank you. As we are describing
% the design of PanDA I thought we should not enter into a too detailed
% description of the DDM design. I eliminated the mistake I had made about
% `centralized data' but left DDM just as a reference. Any better?}

PanDA design supports provenance and traceability for both jobs and data.
Attributes enable provenance by linking jobs and data items, providing
information like ownership or project affiliation. States enable traceability by
providing information about the stage of the execution in which each job or data
item is or has been. Some attributes are assumed to be immutable across
execution and jobs and data items are assumed to be always in one and only one
state.

@@

The implementation of PanDA consists of several interconnected subsystems, most
of them built from off-the-shelf and Open Source components. At a high level,
PanDA's architecture has five main subsystems: PanDA Server, AutoPyFactory,
PanDA Pilot, PanDA Monitoring, and JEDI. Subsystems communicate via dedicated
API or http messaging. Databases are used to store eventful entities like tasks,
jobs and events, and to store information about resources, logs, and accounting.
Each subsystems is implemented by one or more modules as shown in
Figure~\ref{fig:architecture}.

% In the following, we briefly describe the overall architecture of PanDA and
% its components. The description provides all the details required to
% understand how HPC resources have been integrated within PanDA
% (\ref{sec:integration}).

@@

\paragraph{\textbf{PanDA Server}} this subsystem is the central hub service of
PanDA and it operates as a web service~\cite{maeno2011overview}. It runs on
Apache web server, interacting with the back-end database running on separate
servers. It implements the Apache worker model, with many independent processes
handling client requests in parallel. Since all state is maintained in the
central database, the PanDA Server application instance itself is stateless.
Currently, the production version of PanDA utilizes an Oracle database as
backend, but PanDA can also work with MySQL family of databases.

@@

PanDA Server consists of four components implemented in Python:
% and communicating via Python APIs
Task Buffer, Brokerage, Job Dispatcher, and Data Service. Task Buffer manages
PanDA's global queue by: (i) storing jobs into the global queue; (ii) providing
information about and updating the state of each stored job. The Brokerage
component: (i) matches job attributes with site and pilot attributes; (ii)
manages the dispatch of input data to processing sites; and (iii) derives
PanDA's data pre-placement requirement. The Job Dispatcher: (i) receives
requests from active pilots for new jobs to process; (ii) retrieves from Task
Buffer the jobs with highest priority and that can be executed on the site of
the requesting pilot; (iii) creates a wrapper script for the selected job,
tailored to the resource of the requesting pilot; and (iv) monitors the state of
the job execution on the pilot, updating the Task Buffer component. Finally, the
Data Service component: (i) retrieves information from Task Buffer and Brokerage
about the data requirements of each job and the site to which each job has been
assigned; (ii) interacts with the ATLAS DDM to make input files available to the
jobs and manage their output data.

% The job/pilot matching algorithm assigns a weight to each candidate pilot and
% after applying a set of policies, the best ten candidates are taken.

\paragraph{\textbf{PanDA Pilot}} this subsystem implements the execution
environment for PanDA's jobs\cite{nilsson2011atlas}. Implemented in Python, a
PanDA Pilot is launched by a wrapper script submitted to a site. Wrapper scripts
are tailored to the middleware exposed by each site so to maintain PanDA Pilot
site-independent. PanDA Pilot: (i) prepares the execution environment for
PanDA's jobs,
% including  recovering and cleaning up previously failed jobs on that site and
collecting information about the compute, data, and network capabilities of the
site; (ii) requests a job for execution to the Job Dispatcher component of PanDA
Server; (iii) initiates and monitors the execution of pulled jobs, including
setting up the job, transferring its input when required, executing the payload,
transferring output, and checking log files and local space; and (iv) cleans up
when the payload has finished to execute.

@@

% -----------------------------------------------------------------------------
% PILOT DESCRIPTION: COMMENTED OUT BECAUSE WE DESCRIBE IT LATER ON.
% PanDA Pilot implements the pilot paradigm~\cite{turilli2015comprehensive},
% decoupling resource acquisition from job execution via multi-stage scheduling.
% This has several benefits, including increasing throughput by lowering the
% time taken to queue and schedule jobs; enabling both concurrent and
% sequential job execution depending on the number of cores available to the
% pilot and the time for which it remains available; and exposing a unified and
% consistent interface for the scheduling of jobs, independent from the
% middleware of the site on which the pilot has been instantiated.
% -----------------------------------------------------------------------------

@@

Upon bootstrapping, PanDA Pilot downloads major parts of its runtime code via
HTTP from a central Subversion repository. The repository works with an Apache
web server, configured with a memory-based web proxy (Squid). The purpose of the
cache is to reduce the request load on the back-end Subversion server. That
improves performance \mtnote{Do we have data/numbers?} since only occasional
queries trigger a full lookup on the back-end Subversion system, and most
external queries are pulled from memory on the front-end web server. The
benefits of this system include a high-performance code download service,
combined with code updates still being immediately available as soon as they are
committed to source code control.

\paragraph{\textbf{AutoPyFactory}} this subsystem implements a factory to submit
pilots locally and to both grid and cloud
sites~\cite{caballero2012autopyfactory}. Implemented to be executed as a single
demonized process, AutoPyFactory concurrently spawns multiple job submission
processes called APFQueue. Each APFQueue communicates with PanDA Server and the
Brokerage module to collect information about the state of the jobs it manages.
On the basis of this information, an APFQueue submits new pilots to one site.
AutoPyFactory implements functionalities to evaluate the number of pilots
already instantiated on each site, and to apply a consistent and adequate
pressure on their batch queues. Submission is performed via the Condor-G batch
system~\cite{frey2002condor}, enabling submission to multiple type of sites.
AutoPyFactory exposes a dedicated monitoring service, feeding information to the
PanDA Monitoring subsystem. A proxy manager enables certificate-based
authentication and authorization for each site.

@@

\paragraph{\textbf{JEDI}} this subsystem exposes a task abstraction for workload
and workflows specification and interfaces with PanDA Server to manage their
execution on remote sites~\cite{borodin2015scaling}. Implemented as a standalone
service, JEDI partitions tasks to jobs, and binds these jobs to sites via
dedicated queues. JEDI collects information about sites' resource capability,
availability, and performance. Based on this information, JEDI associates sites
with similar capabilities to a queue. Each job is assigned to a queue by
matching the job's requirements to the capability, availability and performance
of the sites of that queue. In this way, JEDI replaces the Brokerage module of
the PanDA Server.
% This module still handles communication between the Task
% Buffer and the Data Service modules.
